---
title: "Visualizing Big Data"
subtitle: "STAE04: Data Visualization"
author: "Johan Larsson"
institute: "The Department of Statistics, Lund University"
date: ""
bibliography: "assets/stae04.bib"
output:
  xaringan::moon_reader:
    css: [default, "assets/stae04.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      titleSlideClass: [middle, center]
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  dev = "png",
  fig.width = 5.2,
  fig.height = 4.2,
  fig.align = "center",
  fig.retina = 3,
  dev.args = list(pointsize = 16),
  crop = TRUE
)

library(tidyverse)
library(kableExtra)

kable <- function(x, ...) {
  knitr::kable(x, escape = FALSE, linesep = "", booktabs = TRUE)
}

library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear",
           style = "markdown",
           dashed = TRUE)
bib <- ReadBib("assets/stae04.bib")

theme_set(theme_bw())

if (!interactive()) {
  theme_set(theme_grey(base_size = 16))
}

options(digits = 3)
```

## Big Data

big data: many variables **and/or** observations

here we consider the case of many observations

the problem with many observations is overlap

.pull-left[
```{r diamonds1, eval = FALSE}
ggplot(diamonds, 
       aes(carat, price)) +
  geom_point()
```
]

.pull-right[
```{r, ref.label = "diamonds1", echo = FALSE, fig.cap = "Carat and price of diamonds."}
```
]

---

## Aesthetics

simple, but not always effective, solutions: modify aesthetics (size, opacity)

.pull-left[
```{r}
ggplot(diamonds, 
       aes(carat, price)) +
  geom_point(alpha = 0.01)  #<<
```
]

.pull-right[
```{r}
ggplot(diamonds, 
       aes(carat, price)) +
  geom_point(cex = 0.01)  #<<
```
]

---

## Facets

sometimes it's enough to split data into facets

```{r, fig.width = 10, fig.height = 5}
ggplot(diamonds, aes(carat, price)) +
  geom_point(alpha = 0.01) +
  facet_wrap(~cut)
```

---

## Bin-Summarize or Smooth

the bin-summarize-smooth framework `r Citep(bib, "wickham2013a")` is a principled
methodology to handle large data

**binning** is the process of grouping values into *bins*; **summarizing** is the
act of computing a summary statistics (count, mean) inside each bean

**smoothing** smooths out noisy data, revealing hard-to-see patterns

we've already encountered binning for one variable: histograms, but they don't work with
two-dimensional data

---

## Binning

simple, but not very sophisticated, idea

```{r, fig.width = 10, fig.height = 2}
trees <- tsibble::as_tsibble(treering)
ggplot(trees, aes(index, value)) + geom_line()
```

--

```{r fig.width = 10, fig.height = 2}
ggplot(trees, aes(index, value)) + geom_line(alpha = 0.2) +
  geom_hline(yintercept = 1, lty = 2) +
  stat_summary_bin(fun = mean, geom = "line", binwidth = 100) #<<
```

---

### Two-Dimensional Binning

a type of two-dimensional histogram, mapping counts to color intead of height
and binning into a two-dimensional geometric shape

```{r, fig.width = 7, fig.height = 4.8}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d(bins = 60) +
  scale_fill_distiller(palette = "Spectral")
```

---

### Hexagonal Binning

rectangles may obscure information for some data and it is therefore usually better
to use hexagons instead, via `geom_hex()`

```{r, fig.width = 7, fig.height = 4.8}
ggplot(diamonds, aes(carat, price)) +
  geom_hex(bins = 60) +
  scale_fill_distiller(palette = "Spectral")
```

---

## Smoothing

smoothing is a powerful transformation for noisy data

smoothing is equivalent to applying a statistical model to the data

```{r, fig.width = 10, fig.height = 4}
ggplot(trees, aes(index, value)) +
  geom_line(alpha = 0.3, size = 0.3) +
  geom_hline(yintercept = 1, lty = 2) +
  geom_smooth() #<<
```

---

### Smoothers are Sensitive

as a thumb rule use only smoothers that you understand

```{r, echo = FALSE}
p <- tsibble::as_tsibble(Nile) %>%
  ggplot(aes(index, value)) +
    geom_line() +
    labs(x = "year", y = "flow")
```

.pull-left[
```{r}
p + geom_smooth(method = "lm")
```
]

--

.pull-right[
```{r}
p + geom_smooth(
  method = "gam",
  formula = y ~ poly(x, 6)
)
```
]

---

### 2D Density Plots

two-dimensional density plots is a type of two-dimensional smoothing

uses a density kernel, which is sensitive to parameters

```{r, fig.width = 7}
ggplot(diamonds, aes(carat, price)) +
  geom_density_2d_filled()
```

---

for other data a density plot may work well

```{r, fig.width = 7, fig.heigt = 5.5}
ggplot(faithful, aes(eruptions, waiting)) +
  geom_density2d_filled() +
  geom_point()
```

hexagonal binning is in general a **safer bet**

---

## References


```{r, results = "asis", echo=FALSE}
PrintBibliography(bib)
```
