---
title: "One-Dimensional Visualizations and Data-Ink"
subtitle: "STAE04: Data Visualization"
author: "Johan Larsson"
institute: "Department of Statistics, Lund University"
date: "`r Sys.Date()`"
bibliography: "assets/stae04.bib"
output:
  xaringan::moon_reader:
    css: [default, "assets/stae04.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      titleSlideClass: [middle, center]
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  dev = "svg",
  fig.width = 3.5,
  fig.height = 3,
  fig.align = "center"
)

library(tidyverse)
library(tactile)
library(lattice)
library(latticeExtra)
library(kableExtra)

kable <- function(x, ...) {
  knitr::kable(x, escape = FALSE, linesep = "", booktabs = TRUE)
}

library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear",
           style = "markdown",
           dashed = TRUE)
bib <- ReadBib("assets/stae04.bib")

theme_set(theme_bw())

if (!interactive()) {
  theme_set(theme_grey(base_size = 14))
}

options(digits = 3)
```

class: center, middle

# One-Dimensional Visualizations

---

## One-Dimensional Visualizations 

.pull-left[
best place to start: continuous or discrete one-dimensional data

often need **statistical transformation** or **position adjustment**
to avoid overlapping

if data is discrete (and not too large), a dotplot (`geom_dotplot()`) is often suitable
]

--

.pull-right[
```{r, fig.height = 2, fig.cap = "Duration between eruptions of the Old Faithful geyser. (The y axis is not useful here and should be removed.)"}
library(tidyverse)

faithful %>%
  ggplot(aes(waiting)) +
  geom_dotplot(binwidth = 1)
```

]

---

## Histograms

.pull-left[
for larger datasets it is often better to
use a statistical transformation

the histogram accomplishes this by *binning* observations to smoothen
the visualization

caution: need to choose how to bin, which may introduce bias
]

.pull-right[
```{r, fig.height = 2.8, fig.cap = "Histogram of Old Faithful data."}
faithful %>%
  ggplot(aes(waiting)) +
  geom_histogram(bins = 15)
```

]

---

## Density Plots

alternative to histograms

attractive for variables that are continuous but sensitive to settings (type of kernel, bandwidth, and more)

.pull-left[
```{r, fig.height = 2}
faithful %>%
  ggplot(aes(waiting)) +
    # tighter bandwidth
    geom_density(bw = 5)
```
]

.pull-right[
```{r, fig.height = 2}
faithful %>%
  ggplot(aes(waiting)) +
    # longer bandwidth
    geom_density(bw = 10)
```
]

often good idea to add a **rug** layer (`geom_rug()`) to density plots

---

## Boxplots

.pull-left[
most common type:

  - **median** as middle bar, 
  - 1st and 3rd **quartiles** as edges of box,
  - whiskers to last observation within 1.5 times the
    inter-quartile range, and
  - points (or starts) beyond whiskers

typically useful when there are categories in the plot

more abstraction (aggregation) than histogram or densityplot
]

.pull-right[
```{r, fig.height = 2, fig.cap = "Boxplot of the Old Faithful data."}
faithful %>%
  ggplot(aes(waiting)) +
    geom_boxplot()
```
]

---

## Violin Plots

a type of density plot, often used instead of boxplot

same pitfalls as with other density plots

```{r, fig.width = 5, fig.height = 2.5, fig.cap = "geom_violin() does not work with a single variable, so we need y = 1 here."}
faithful %>%
  ggplot(aes(waiting, y = 1)) +
    geom_violin()
```

---

class: center, middle

# Data-Ink

---

## Data-Ink

always what you are doing with your **ink** in a visualization

> Above all else show the data (Tufte, 1983)

data-ink: ink used to draw data

$$\text{data-ink ratio} = \frac{\text{data-ink}}{\text{total ink used in visualization}}$$
--

.pull-left[
```{r, echo = FALSE, fig.cap = "low data-ink ratio", fig.height = 2.5}
drop_na(msleep, sleep_total) %>%
  group_by(vore) %>%
  summarize(m_sleep = mean(sleep_total)) %>%
  ggplot(aes(vore, m_sleep)) +
  geom_col() +
  ylab("mean sleep")
```
]

.pull-right[
```{r, echo = FALSE, fig.cap = "high data-ink ratio", fig.height = 2.5}
msleep %>%
  ggplot(aes(vore, sleep_total)) +
  geom_boxplot()
```
]

---

## Edward Tufte

Edward Tufte's work has been seminal in dataviz

has developed variations on standard plots (some available in **ggthemes**)

.pull-left[
```{r}
library(ggthemes)
ggplot(mpg, aes(drv, cty)) +
  geom_tufteboxplot() +
  theme_tufte()
```
]

.pull-right[
```{r}
ggplot(mpg, aes(drv, cty)) +
  geom_boxplot(
    stat = "fivenumber"
  )
```
]
