---
title: "Two-Dimensional Visualizations"
subtitle: "STAE04: Data Visualization"
author: "Johan Larsson"
institute: "The Department of Statistics, Lund University"
date: ""
bibliography: "assets/stae04.bib"
output:
  xaringan::moon_reader:
    css: [default, "assets/stae04.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      titleSlideClass: [middle, center]
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  dev = "svg",
  fig.width = 3.5,
  fig.height = 3,
  fig.align = "center"
)

library(tidyverse)
library(tactile)
library(lattice)
library(latticeExtra)
library(kableExtra)

kable <- function(x, ...) {
  knitr::kable(x, escape = FALSE, linesep = "", booktabs = TRUE)
}

library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear",
           style = "markdown",
           dashed = TRUE)
bib <- ReadBib("assets/stae04.bib")

theme_set(theme_bw())

if (!interactive()) {
  theme_set(theme_grey(base_size = 14))
}

options(digits = 3)
```

## Associations

.pull-left[
two-dimensional visualizations show **associations**: how does $x$ relate to $y$?

visualizing two dimensions is usually **straightforward**, particularly when both
variables are continuous

scatter plots (`geom_point()`) are the bread and butter of visualizations of
two continuous variables

```{r, eval = FALSE}
library(tidyverse)
ggplot(swiss, 
       aes(Examination, 
           Fertility)) +
  geom_point()
```

]

.pull-right[
```{r, fig.cap = "Fertility and % of draftees receiving highest mark on army examination in French-speaking provinces in Switzerland.", echo = FALSE}
library(tidyverse)
ggplot(swiss, 
       aes(Examination, 
           Fertility)) +
  geom_point()
```
]

---

## Patterns

understanding patterns is key to being able to visualize data effectively

```{r, fig.width = 5, fig.height = 5, echo = FALSE}
set.seed(1453)

n <- 100
x <- runif(n, 1, 6)

y_clust <- double(n)
y_clust[x <= 3] <- rnorm(n, -4)
y_clust[x > 3] <- rnorm(n, 4)

d <- tibble(
  x = x,
  noise = runif(n),
  linear = x + rnorm(n, 0, 1),
  nonlinear = 2^x + 0.5*x*rnorm(n, 0, 5),
  clusters = y_clust
)

d_long <- pivot_longer(d, -x) %>%
  mutate(name = fct_relevel(name, "noise", "linear", "nonlinear", "clusters"))

ggplot(d_long, aes(x, value, group = name)) +
  geom_point() +
  facet_wrap(~name, scales = "free") +
  xlab(NULL)
```

---

## Variation

.pull-left-40[
a good visualization should describe the variation in the data
]

.pull-right-60[
```{r, echo = FALSE, fig.height = 2.6, fig.width = 4.5}
d <- tibble(x = 1:100,
            low_variance = x + rnorm(100, 0, 4),
            high_variance = x + rnorm(100, 0, 20))

pivot_longer(d, -x) %>%
  ggplot(aes(x, value, group = name)) +
  geom_point() +
  ylab("y") +
  facet_wrap(~name)
```
]

--

.pull-left-60[
```{r, echo = FALSE, fig.height = 2.6, fig.width = 4.5}
d <- tibble(x = 1:100,
            y = rnorm(100, 0, x))

ggplot(d, aes(x, y)) +
  geom_point()
```
]

.pull-right-40[
sometimes variation has a pattern of itself
]



---

## Time

time has a specific order, so the line geom is often natural

always place time on the horizontal (x) axis

```{r, fig.cap = "Median unemployment duration in the US.", fig.width = 6, fig.height = 2.7}
ggplot(economics, aes(date, uempmed)) +
  geom_line() +
  xlab("Time") +
  ylab("Unemployment duration")
```

---

## Geoms and Time

often useful to use points and lines together when measurements
are taken at irregular intervals or seldom

```{r, fig.cap = "Weight of a chick over time.", fig.width = 4}
filter(ChickWeight, Chick == 5) %>%
  ggplot(aes(Time, weight)) +
  geom_point() +
  geom_line()
```


---

## Categorical Variables

categorical variables (factors) can of course be used
as the second dimension

```{r, fig.cap = "Boxplot of total sleep duration for different mammals."}
ggplot(msleep, aes(vore, sleep_total)) +
  geom_boxplot()
```

---

## Ordered Factors

pay attention to whether or not the factor is **ordered**

.pull-left[
```{r, echo = FALSE, fig.height = 4}
mutate(OrchardSprays, treatment = fct_shuffle(treatment)) %>%
  ggplot(aes(treatment, decrease)) +
  geom_boxplot()
```
]

--

.pull-right[
```{r, echo = FALSE, fig.height = 4}
ggplot(OrchardSprays, aes(treatment, decrease)) +
  geom_boxplot()
```
]

---

## Discrete Data

a problem when one (or both) variables are discrete (or categorical)
is that they might overlap

a simply solution is to use opacity (`alpha`)

.pull-left[
```{r}
ggplot(mpg, aes(cty, hwy)) +
  geom_point() 
```
]

--

.pull-right[
```{r}
ggplot(mpg, aes(cty, hwy)) +
  geom_point(alpha = 0.2) 
```
]

---

## Scales

for some variables, it is more natural to study them on a different scale

other times it is simply convenient (more compact)

.pull-left[
```{r, fig.height = 2.7}
ggplot(msleep, 
       aes(brainwt, 
           sleep_total)) +
  geom_point() +
  scale_y_continuous()
```
]

--

.pull-right[
```{r, fig.height = 2.7}
ggplot(msleep, 
       aes(brainwt, 
           sleep_total)) +
  geom_point() +
  scale_x_log10()
```
]









