---
title: "Assignment and Project Instructions"
author: "Johan Larsson"
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.width = 4,
                      fig.height = 3.6)
```

## Instructions

For each assignment, as well as for the project, you need to hand in a `.pdf` report. 
This report should be a written 
response to the tasks given in the assignment. It should respond to all the questions in 
the assignment in a **clear** and **succinct** manner.

### Format

We **strongly** recommend that you use R Markdown to author your report. If you do, please use the [submission template](https://raw.githubusercontent.com/stat-lu/STAE04/master/stae04-template.Rmd)
that we have provided for you. To use the template, go to the link, and either

- right click anywhere on the page, select *save page as* to save the `.Rmd` file to
  your hard drive, and open the file in R Studio, or
- click `File -> New File -> R Markdown -> Create Empty Document` and copy-paste the 
  contents of the [page](https://raw.githubusercontent.com/stat-lu/STAE04/master/stae04-template.Rmd) 
  into the editor.

This template also serves as an introduction to R Markdown. To get started, simply open the document in R Studio and read through the code.

If you prefer to use a WYSIWYG (what you see is what you get) editor such as Libre Office, Google Docs, or Microsoft Word, you will likely want to either use `ggplot2::ggsave()` or the built-in export utility in the plot pane of R Studio to save your plots to your hard drive and then input them manually into your document. Here is a simple example of how to do this:

```{r, eval = FALSE}
library(ggplot2)

ggplot(mpg, aes(cty, hwy)) +
  geom_point()

ggsave("filename.png", width = 3, height = 2.3)
```

If you choose to work with a WYSIWYG editor, we recommend that you use a single `.R` file for each assignment. If you use R Markdown, you will only need a `.Rmd` file for each assignment.

### Detailed Requirements

Please pay attention to the following:

- Your report does not need to be verbose but it must be **clear** and **coherent**. Use full
  sentences and describe the steps you take and code you've included.
- Any figures you include should have descriptive labels and captions (use `fig.cap` in
  your code chunk headers if you're using R Markdown). See [The Anatomy of a Plot](https://canvas.education.lu.se/courses/5769/pages/the-anatomy-of-a-plot) for
  examples and more instruction on how to format your figures.
- Only include code when you are **instructed** to do so. When you do include code, please
  avoid having it escape into the margin by using multiple lines to separate code. 
  Good code style is not a requirement in this course but is nevertheless good practice.
  It might be a good idea to conform to a style guide such as
  [the tidyverse style guide](https://style.tidyverse.org/).
  
### Peer Review and Examination

For each assignment you will need to peer review **three other reports**. To complete the reviews, you need to fill out the **rubric** associated with the assignment and add a general comment on the submission. In addition, we also encourage you to provide **annotated feedback** using the integrated tools in Canvas. Please see [this page](https://community.canvaslms.com/t5/Student-Guide/How-do-I-submit-a-peer-review-to-an-assignment/ta-p/293) to read more about peer-reviewing in canvas.

Finally, please remember to provide **constructive** feedback.

#### You have one week to complete your peer review!

If you have not completed your peer review within one week past the deadline, you will **fail the assignment**. There is unfortunately 
no way in Canvas to create a deadline for the peer reviews, so you have to keep track
of this deadline manually.

#### Grading

The assignments and project will be graded by an instructor or teaching assistant as **pass** or **fail**. To pass the course, **you need to pass each assignment**. If you fail any assignment (or the project), you will get an additional chance to hand in these after the conclusion of the course.

### Hints

- Functions in the tidyverse are well-documented. Students who are successful at these
  assignments often make good use of the help files for the various function (in RStudio
  simply place the cursor on the function name and hit F1).
- To run code in R Studio inside a R Markdown document, highlight the code and
  hit `CTRL/CMD+Enter`. (There is no need to copy it directly into the terminal.)
- To create a code chunk in your R Markdown document, use the shortcut `CTRL/CMD+ALT+I`.
- If you want code in your R Markdown document to appear inside your final report, use
  `echo = TRUE` in the header of the code chunk.
- The size of figures can be controlled using the `fig.width` and `fig.height` options
  in the code header (if you use R Markdown).
