---
title: "R Markdown Template"
subtitle: "STAE04: Data Visualization"
papersize: a4
author: "Johan Larsson"
affiliation: "Department of Statistics, Lund University"
date: "`r Sys.Date()`"
toc: true
header: true
graphics: true
colorlinks: true
output: 
  rmarkdown::pdf_document:
    template: NULL
    number_sections: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 2.5,
  fig.height = 2.2
)

library(tidyverse)

# set ggtheme options
theme_set(theme_grey(base_size = 8))

# some sensible options
options(
  digits = 3, # limit number of significant digits
  width = 63  # limit width of code
)
```

# Rmarkdown

This document is written using [R Markdown](http://rmarkdown.rstudio.com). R Markdown
is a syntax for formatting documents that lets you focus on the 
content instead of formatting. You write text, including code, in Rmarkdown syntax
and then Rmarkdown (plus a handful of very useful packages) turn that text into
a neatly formatted document.

This document serves as both a template for your submissions during the data visualization
course as well as an introduction into R Markdown. To follow along, you should load
this file in RStudio. To get started, you need to install the **rmarkdown** package.

```{r, eval = FALSE}
install.packages("rmarkdown")
```


A Rmarkdown file ends with `.Rmd` and is best edited using RStudio. Each Rmarkdown file
starts with a so called YAML block; here is a bare-bones one:

```yaml
---
title: "An Awesome Title"
author: "Fantastic Me"
date: "2020-09-28"
output: pdf_document
---
```

For this course, you will not need to
change anything in the YAML block other than the `author`, `date`, and `title` fields.

## Text Formatting

Rmarkdown uses [Pandoc markdown](https://pandoc.org/MANUAL.html#pandocs-markdown)
for formatting text, which uses a special---but very simple---syntax for formatting
text. To write bold text for instance, you use **two stars** to surround text. To
write italic text, you use a *single star*.

Links can be specified using square brackets
with the text for the link followed by parantheses for the
actual link, such as [this link to pandoc manual](https://pandoc.org/MANUAL).

Section titles are written using hash tags, as in this document. One hash tag
indicates the highest-level section heading, two the second-highest, and so on.

## Code Chunks

What makes RMarkdown special is that is allows us to write code chunks containing
R code. This is a code chunk that plots a simple plot using ggplot2:

```{r fig.cap = "A Caption."}
library(tidyverse)

ggplot(msleep, aes(brainwt, sleep_rem)) +
  geom_point() + 
  scale_x_log10() +
  labs(x = "Brain Weight (lbs)", y = "REM Sleep (hours)")
```

You can control many settings by changing the parameters in the headline of the
code chunk. The following chunk changes the width and height of the figure, which
are settings that will be very useful to you.

```{r fig.cap = "A Caption.", fig.width = 4, fig.height = 1.5}
ggplot(msleep, aes(brainwt, sleep_rem)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "Brain Weight (kg)", y = "REM Sleep (hours)")
```

## Rendering an Rmarkdown Document

When you are done with your writing, you can render the final document by *knitting* it.
In RStudio, the shortcut for this is `CTRL+Shift+K`. When a document is knitted, all your
code in the document is first run and then the result is passed on to pandoc, which
produces the final document for you.

R Markdown files can be knitted to a wide range of different medium, including
PDF files, HTML, and word documents. In this course, however, you will always
produce PDF files from your R Markdown documents.

## Reproducibility

Authoring your documents using R Markdown facilitates reproducibility. Because you
need to supply all the code used to produce your paper in the `.Rmd` file, this
makes it much easier for other people to re-run your analysis and use your code.

# Submissions

In this course, you will use R Markdown for all your assignment submissions. For each
of these you will need to submit both the `.Rmd` source file as well
as the knitted `.pdf` file. Documents from office packages (such as Microsoft Office)
will not be accepted.

# Learning More About R Markdown

If you want to learn more about R Markdown, a good place to start is
at <https://rmarkdown.rstudio.com/lesson-1.html>.
